---
title: "Halloween Candy Mini Project"
author: "Patric Young"
format: pdf
editor: visual
---

# 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"
candy = read.csv("candy-data.csv", row.names=1)
head(candy)
```

-   **Q1**. How many different candy types are in this dataset?

    ```{r}
    nrow(candy)
    ```

    There are 85 different candy types in this data set.

-   **Q2**. How many fruity candy types are in the data set?

    ```{r}
    table(candy$fruit)
    ```

There are 38 fruity candy types are in this data set.

# 2. What is your favorate candy?

-   **Q3**. What is your favorite candy in the data set and what is it's `winpercent` value?

```{r}
candy["Twix", ]$winpercent
```

Favorite candy: Twix

Winpercent: 81.64%

-   **Q4**. What is the `winpercent` value for "Kit Kat"?

    ```{r}
    candy["Kit Kat", ]$winpercent
    ```

    76.77%

-   **Q5**. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

49.65%

```{r}
library("skimr")
skim(candy)

```

-   **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

    The winpercent variable is on a different scale to the majority of the other columns in the data set. While the other variables are on a scale of 0 to 1, winpercent is on a 0 to 100 percent scale.

-   **Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

    A zero represents no "vote" was casted for Chocolate whereas a one represents that a vote was casted.

-   **Q8**. Plot a histogram of `winpercent` values

```{r}
library("ggplot2")

```

```{r}
ggplot(candy, aes(x = winpercent)) +
geom_histogram()
```

-   **Q9**. Is the distribution of `winpercent` values symmetrical?

    The distribution of `winpercent` values are not symmetrical.

-   **Q10**. Is the center of the distribution above or below 50%?

The center of distribution is slightly below 50%

-   **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

    On average, Chocolate candy is ranked higher than fruit candy

-   **Q12**. Is this difference statistically significant?

    Since the p-value falls below the significance level, the result is statistically significant.

# 3. Overall Candy Rankings

-   **Q13**. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

-   **Q14**. What are the top 5 all time favorite candy types out of this set?

The top 5 all time favorite candy types out of this set are Reeses Peanut Butter Cup, Reeses Minatures, Twix, Kit Kat, and Reeses Pieces.

**Q15**. Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

-   **Q16**. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

**Q17**. What is the worst ranked chocolate candy?

The worst ranked chocolate candy are Sixlets.

**Q18**. What is the best ranked fruity candy?

The best ranked fruity candy are Starbursts.

# 4. Taking a look at pricepercent

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

**Q19**. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

Hershey's Krackel is the highest `winpercent` candy (62.28%) with the lowest `pricepercent`(0.918%).

**Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candy types in the data set are Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular of these are the Nik L Nips.

-   **Q21**. Make a bar plot again with `geom_col()` this time using `pricepercent` and then improve this step by step, first ordering the x-axis by value and finally making a so called "dot chat" or "lollipop" chart by swapping `geom_col()` for `geom_point()` + `geom_segment()`.

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

**Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity candies are the two most anti-correlated variables.

**Q23**. Similarly, what two variables are most positively correlated?

Chocolate and Chocolate are the two most positively correlated variables, followed by Chocolate and Bar.

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE )
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

-   **Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

    Fruity, Hard, and Pluribus are the 3 variables picked up strongly by PC1 in the positive direction. They make sense to me because it means these 3 variables are correlated just like Chocolate and Bar.
